---
title: "ddR README"
author: "Edward Ma, Indrajit Roy"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

ddR is both an API and an R package that permits the declaration of 'distributed' objects (i.e., `dlist`, `dframe`, `darray`), and facilitates parallel operations on these data structures using R-style `apply` functions. It also allows different backends (that support ddR, and have ddR "driver" written for them), to be dynamically activated in the R user's environment, to be selected for use with the API.

To get started, simply load the library and select a backend to use. If you'd like to use a custom backend that isn't `Parallel`, you'll also need to load the library or code containing the driver for that backend. (Below, you'd also run the lines that are commented out)

```{r}
library(ddR)
## Run the next two lines also to use the Distributed R backend
# library(distributedR.ddR)
# useBackend(distributedR)
```

### Creating Distributed Objects

There are two ways to create distributed objects in ddR.

1. Using the constructor functions.
2. Using `dmapply` and/or `dlapply` (`dlists` only).

## Using the constructor functions

A `dlist` may be created using the constructor with a comma-separated list of arguments. For example:

```{r}
my.dlist <- dlist(1,2,3,4,5)
my.dlist
```

Note that this printout shows you a lot of useful information, including metadata about the object, partition sizes, etc. There are 5 partitions in `my.dlist`, because by default, the constructor creates as many partitions as there are arugments.

However, you may also specify the partitioning directly using the constructor:

```{r}
my.dlist <- dlist(1,2,3,4,5,nparts=3)
my.dlist
```

When `nparts` is supplied, it will create that many partitions in the output object, with a best effort splitting of data between those partitions. In this case, since 5 isn't divisible by 3, it divided the data into lengths of 2, 2, and 1.

The constructors for `darray` and `dframe` are different. For example, you might initialize a `darray` in the following manner:

```{r}
my.darray <- darray(dim=c(4,4),psize=c(2,2),data=3)
my.darray
```

This makes `my.darray` a `darray` that's filled with 3s. Each partition is of size 2x2, and the dimensions of `my.darray` are 4x4. For `dframe`, the constructor follows the same format.

## Using `dmapply` or `dlapply`

Constructors are handy when you want to initialize distributed objects quickly, but as `ddR` is a functional-programming API, every operation you do will generate new objects. This means that the primary way in which you will create distributed objects will be via `dmapply` and `dlapply`.

Here's how we would create the same `dlist` that we created above using `dlapply`:

```{r}
my.dlist <- dlapply(1:5,function(x) x)
my.dlist
```

What did I do? I ran the distributed `lapply` function, `dlapply`, on the function argument vector `1:5`, assigning to each element the value of the vector for that iteration. To specify `nparts`, I could also supply `nparts`, just as I did in the constructor function:

```{r}
my.dlist <- dlapply(1:5,function(x) x, nparts=3)
my.dlist
```

The API with `dmapply` for `darray` and `dframe` is slightly more complex, though not substantially so. When creating these data structures using `dmapply`, the API needs to know some information, such as how to partition the output in 2d-manner, as well as how to combine intermediate results of dmapply within each partition. Therefore, you need to also supply arguments for the following parameters:

1. `output.type`, as either `"darray"` or `"dframe"` (default is `"dlist"`).
2. `nparts`. Instead of just a scalar value, this needs to be a vector of length 2, in order to specify how to two-dimensionally partition the output `darray` or `dframe`. For example, `nparts=c(2,2)` means you want the four partitions resulting from `dmapply` to be stitched together in a 2 by 2 fashion.
3. `combine` This is needed if you ever have more than one `dmapply` iteration per partition. For example, you may `dmapply` on arguments of length 10, when you only have 4 partitions in your output `darray` or `dframe`. When that happens, `combine` allows you to specify how you'd like to combine this data. You may like to think of this operation as what is called within each partition together on the results, before the aggregate data of the partition assumes its form. `combine` can be either `c` (default), `rbind`, or `cbind`.

So, let's create a 4x4 `darray`, consisting of 4 2x2 partitions, where each partition contains values equal to its partition id. To do that, we can do the following:

```{r}
my.darray2 <- dmapply(function(x) matrix(x,2,2), 1:4, output.type="darray", combine="rbind", nparts=c(2,2))
my.darray2
```

Even though we didn't `rbind` anything, as each iteration of `dmapply` was one partition of the result, the `combine` value was necessary, since the default value of `combine` is `c`, which flattens and vectorizes the results within each partition (this is the default behavior of R's `mapply`). So `rbind` prevents this from happening, and the matrix structure is retained. We can look at what's stored in `my.darray2` by using the `collect` operator, which brings the data from the distributed backend to the local R instance, as a local R object:

```{r}
my.array <- collect(my.darray2)
my.array
```

### Collect, parts

To-be-completed

### Repartitioning

To-be-completed

### More Examples

Init'ing a dlist:
```{r}
a <- dmapply(function(x) { x }, rep(3,5))
collect(a)
```

Printing `a`:
```{r}
a
```

`a` is now a distributed object in ddR. Note that we did not specify the number of partitions of the output, but by default it went to the length of the inputs (5). If we wanted to specify how the output should be partitioned, we can use the `nparts` parameter to `dmapply`:

Adding 1 to first element of `a`, 2 to the second, etc.

```{r}
b <- dmapply(function(x,y) { x + y }, a, 1:5,nparts=1)
b
```

As you can see, `b` only has one partition of 5 elements.

```{r}
collect(b)
```
Some other operations:
`

Adding `a` to `b`, then subtracting a constant value
```{r}
addThenSubtract <- function(x,y,z) {
  x + y - z
}
c <- dmapply(addThenSubtract,a,b,MoreArgs=list(z=5))
collect(c)
```

Accessing dobjects by parts:

```{r}
d <- dmapply(function(x) length(x),parts(a))
collect(d)
```

We partitioned `a` with 5 parts and it had 5 elements, so the length of each partition is of course 1.

However, `b` only had one partition, so that one partition should be of length 5:

```{r}
e <- dmapply(function(x) length(x),parts(b))
collect(e)
```

Note that `parts()` and non-parts arguments can be used in any combination to dmapply. `parts(dobj)` returns a list of the partitions of that dobject, which can be passed into dmapply like any other list. `parts(dobj,index)`, where `index` is a list, vector, or scalar, returns a specific partition or range of partitions of `dobj`.

We also have support for `darrays` and `dframes`. Their APIs are a bit more complex, and this guide will be updated shortly with that content.

For a more detailed example, you may view (and run) the example scripts under /examples.

## Using the Distributed R backend

Use the Distributed R library for ddR:
```r
library(distributedR.ddR)
```

```
## Loading required package: distributedR
## Loading required package: Rcpp
## Loading required package: RInside
## Loading required package: XML
## Loading required package: ddR
## 
## Attaching package: 'ddR'
## 
## The following objects are masked from 'package:distributedR':
## 
##     darray, dframe, dlist, is.dlist
```

```r
useBackend(distributedR)
```

```
## Master address:port - 127.0.0.1:50000
```

Now you can try the different list examples which were used with the 'parallel' backend.

## How to Contribute

You can help us in different ways:

1. Reporting [issues](https://github.com/vertica/ddR/issues).
2. Contributing code and sending a [Pull Request](https://github.com/vertica/ddR/pulls).

In order to contribute the code base of this project, you must agree to the Developer Certificate of Origin (DCO) 1.1 for this project under GPLv2+:

    By making a contribution to this project, I certify that:
    
    (a) The contribution was created in whole or in part by me and I have the 
        right to submit it under the open source license indicated in the file; or
    (b) The contribution is based upon previous work that, to the best of my 
        knowledge, is covered under an appropriate open source license and I 
        have the right under that license to submit that work with modifications, 
        whether created in whole or in part by me, under the same open source 
        license (unless I am permitted to submit under a different license), 
        as indicated in the file; or
    (c) The contribution was provided directly to me by some other person who 
        certified (a), (b) or (c) and I have not modified it.
    (d) I understand and agree that this project and the contribution are public and
        that a record of the contribution (including all personal information I submit 
        with it, including my sign-off) is maintained indefinitely and may be 
        redistributed consistent with this project or the open source license(s) involved.

To indicate acceptance of the DCO you need to add a `Signed-off-by` line to every commit. E.g.:

    Signed-off-by: John Doe <john.doe@hisdomain.com>

To automatically add that line use the `-s` switch when running `git commit`:

    $ git commit -s
